Efficiency

### 2017-12-29 DraftExec.Rmd
Big Push, need to migrate

@@

@@

- id: hamilton1997
  title: Jet Torque
  author:
  - family: Waterjets
    given: Hamilton
  volume: 8
  URL: ''
  publisher: Hamilton Waterjets
  issued:
    year: 1997
    month: 3
---

@@

### Conservation of Energy

@@

#### Venturi Effect and Flow Diffusion
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect. The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.  This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].

#### Energy waste and efficiency
The conservation of energy also has useful implications for how the efficiency of systems is thought about.  In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.  A perfectly efficient system won't do any work that isn't for the intended purpose.  In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.  Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.  Noise in the water also reflects energy which is wasted.

### Boundary Layer
When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.  At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.  The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.  Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.

### Turbulent and Laminar Flow

### Pressure, Thrust, and Momentum

### Cavitation

### Flow separation

# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations

# Propulsive performance of the pumpjet, or decelerating ducted propeller

## Efficiency Relationships, and various curves

## Practical and recent evidence of efficiency relationships

>>>>

- id: fujisawa1995
  title: Measurement of Basic Performances for Waterjet Propulsion Systems in Water Tunnel
  author:
  - family: Fujisawa
    given: Nobuyuki
  volume: 2
  URL: 'downloads.hindawi.com/journals/ijrm/1995/936068.pdf'
  publisher: International Journal of Rotating Machinery
  type: article-journal
  issued:
    year: 1995

@@

- id: NPdotnet2017
  title: Laminar Flow
  author:
  - family: 
    given: 
  URL: 'http://www.nuclear-power.net/nuclear-engineering/fluid-dynamics/laminar-flow-viscous/'
  publisher: nuclear-power.net
  issued:
    year: 2017

@@

- id: patrick2017
  title: Senate Estimates Committee 
  author:
  - family: Patrick
    given: Rex
  URL: 'http://parlview.aph.gov.au/mediaPlayer.php?videoID=379261&operation_mode=parlview#/4%20from%20around%201201'
  publisher: Australian Parliament House
  issued:
    year: 2017
    month: 10

- id: hamilton1997
  title: Jet Torque
  author:
  - family: Waterjets
    given: Hamilton
  volume: 8
  URL: 'http://www.hamiltonmarine.co.nz/includes/files_cms/file/JetTorque%2008.pdf'
  publisher: Hamilton Waterjets
  issued:
    year: 1997
    month: 3

- id: mollard2011
  title: Ship Resistance and Propulsion
  author:
  - family: Mollard
    given: Anthony F.
  author:
  - family: Turnock
    given: Stephen R.
  author:
  - family: Hudson
    given: Dominic A.
  URL: 'http://dl.kashti.ir/ENBOOKS/NEW/Ship%20resistance%20and%20propulsion%20_%20F.Molland%20(2011).pdf'
  publisher: Cambridge University Press
  issued:
    year: 2011

- id: shin2015
  title: Methods for Cavitation Prediction on Tip-Modified Propellers in Ship Wake Fields
  author:
  - family: Shin
    given: Keun Woo
  - family: Regener
    given: Pelle Bo
  - family: Andersen
    given: Poul
  URL: 'http://www.marinepropulsors.com/proceedings/2015/WA2-1.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: moulijn2015
  title: Application of Various Computational Methods for Predict the Performance of Cavitation of Ducted Propellers
  author:
  - family: Moulijn
    given: Joost
  URL: 'www.marinepropulsors.com/proceedings/2015/WB3-1.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: lewis1988
  title: Principals of Naval Architecture
  author:
  - family: Van Manen
    given: J.D.
  - family: Van Oossanen
    given: P.
  volume: 2
  URL: 'http://opac.vimaru.edu.vn/edata/EBook/Principles%20of%20Naval%20achitecture%202.pdf'
  publisher: The Society of Naval Architects and Marine Engineers
  editor: 
  - family: Lewis
    given: Edward V.
  issued:
    year: 1988

- id: li2013
  title: Dynamic Characteristics of Rotating Stall in Mixed Flow Pump
  author:
  - family: Li
    given: Xiaojun
  - family: Yuan
    given: Shouqi
  - family: Pan
    given: Zhongyong
  - family: Li
    given: Yi
  - family: Liu
    given: Wei
  URL: 'https://www.hindawi.com/journals/jam/2013/104629/'
  publisher: Hindawi Publishing Corporation Journal of Applied Mathematics
  issued:
    year: 2013
    month: 8

- id: kamal2015
  title: Powering for Medium Speed Wave-Piercing Catamarans comparing Waterjet and Screw Propeller Performance using Model Testing
  author:
  - family: Kamal
    given: Iwan Mustaffa
  - family: Zürcher
    given: Konrad
  - family: Bose
    given: Neil
  - family: Binns
    given: Jonathan
  - family: Chai
    given: Shuhong
  - family: Davidson
    given: Gary
  URL: 'http://www.marinepropulsors.com/proceedings/2015/MA3-2.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: giles2010
  title: The Advanced WaterJet Propulsor Performance and Effect on Ship Design
  author:
  - family: Giles
    given: Will
  - family: Dinham-Peren
    given: Tom
  - family: Amaratunga
    given: Shane
  - family: Vrijdag
    given: Arthur
  - family: Partridge
    given: Richard
  URL: 'https://www.bmtdsl.co.uk/media/6097875/BMTDSL-The-Advanced-Waterjet-Confpaper-INEC-May10.pdf'
  publisher: Tenth International Naval Engineering Conference
  issued:
    year: 2010
    month: 5

- id: haimov2010
  title: Ducted Propellers. A Solution for Better Propulsion of Ships. Calculations and Practice.
  author:
  - family: Haimov
    given: H.
  - family: Bobo
    given: M.J.
  - family: Vicario
    given: J.
  - family: del Corral
    given: J.
  URL: 'http://www.cehipar.es/__files/users/publicaciones/217.pdf'
  publisher: Ministerio de Defensa
  issued:
    year: 2010
    month: 5

- id: abdel2010
  title: Design of a Multi-Component Propulsor
  author:
  - family: Abdel-Maksoud
    given: Moustafa
  - family: Steden
    given: Max
  - family: Hundemer
    given: Jochen
  publisher: 28th Symposium of Naval Hydrodynamics
  issued:
    year: 2010
    month: 9

- id: bakker2006
  title: Boundary Layers and Separation
  author:
  - family: Bakker
  URL: 'http://www.bakker.org/dartmouth06/engs150/11-bl.pdf'
  publisher: Dartmouth College
  issued:
    year: 2002

- id: mcbride1979
  title: Design and analysis of turbomachinery in an incompressible, steady flow using the streamline curvature method
  author:
  - family: McBride
    given: Mark W.
  URL: 'www.dtic.mil/dtic/tr/fulltext/u2/a071107.pdf'
  publisher: Naval Sea Systems Command
  issued:
    year: 1979
    month: 7

- id: wislicenus1986
  title: Preliminary Design of Turbopumps and Related Machinery
  author:
  - family: Wislicenus
    given: George F.
  URL: 'https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/19870008232.pdf'
  publisher: NASA
  issued:
    year: 1986
    month: 10

- id: wislicenus1973
  title: Hydrodynamic Design Principles of Pumps and Ducting for Waterjet Propulsion
  author:
  - family: Wislicenus
    given: George F.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/775620.pdf'
  publisher: Naval Ship Research & Development Center
  issued:
    year: 1973
    month: 6

- id: brandau1967
  title: Apects of performance evaluation of waterjet propulsion systems and a critical review of the state-of-the-art
  author:
  - family: Brandau
    given: John H.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/827069.pdf'
  publisher: Naval Ship Research & Development Center
  issued:
    year: 1967
    month: 10

- id: henderson1964
  title: A Method for the Design of Pumpjets
  author:
  - family: Henderson
    given: R.E.
  - family: McMahon
    given: J.F.
  - family: Wislicenus
    given: G.F.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/439631.pdf'
  publisher: Ordinance Research Laboratory, Department of Navy
  issued:
    year: 1964
    month: 5

- id: MAN2017
  title: Basic Principals of Ship Propulsion
  author:
  - family: MAN Diesel & Turbo
  URL: 'https://marine.man.eu/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0'
  publisher: MAN Diesel & Turbo
  issued:
    year: 2017

- id: oosterveld1970
  title: Wake Adapted Ducted Propellers
  author:
  - family: Oosterveld
    given: IR.M.W.C.
  URL: 'http://www.marin.nl/web/Publications/Publication-items/Wake-Adapted-Ducted-Propellers.htm'
  publisher: Netherlands Ship Model Basin Wageningen - Netherlands
  issued:
    year: 1970

- id: davies2017
  title: Pump up the pump jet?
  author:
  - family: Davies
    given: Andrew
  URL: 'https://www.aspistrategist.org.au/pump-up-the-pump-jet/'
  publisher: Australian Strategic Policy Institute
  issued:
    year: 2017
    month: 11

- id: stanford2017
  title: Australia's Future Submarine Getting This Key Capability Right
  author:
  - family: Jon
    given: Stanford
  URL: 'http://www.insighteconomics.com.au/reports/2017_Insight_Economics_Submarine_Report.pdf'
  publisher: Insight Economics
  issued:
    year: 2017
    month: 09

---

@@

## Conservation of Energy

@@

### Venturi Effect 
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect.

### Flow Diffusion
The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.  This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].

### Energy waste and efficiency
The conservation of energy also has useful implications for how the efficiency of systems is thought about.  In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.  A perfectly efficient system won't do any work that isn't for the intended purpose.  In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.  Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.  Noise in the water also reflects energy which is wasted.

## Boundary Layer
When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.  At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.  The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.  Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.  In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown.

![The development of a boundary layer as shown in @mollard2011](/Users/aidanmorrison/submarines/BoundaryLayer.png){width=500px}

## Turbulent and Laminar Flow
The way that fluids move relative to a surface can one of either two methods.  In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.

The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.  There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.  Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.  Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.

![A simple comparison of laminar and turbulent flows from @NPdotnet2017](/Users/aidanmorrison/submarines/LaminarTurbulent.png)

The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.  A fuller discussion of these important concepts can be found in [@NPdotnet2017].

Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.  Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.  Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.

In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.  Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise. Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.

## Cavitation
Cavitation is the rapid expansion and collapse of a void, or bubble in water. Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).

Imagine what happens when something moves very fast through water.  The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.  If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.  (This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.  Soon the water catches up, and the bubble implodes with a pop, leaving only tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.  Whilst some energy turns to heat (the remaining steam in the tiny bubbles) quite a bit is propagated away as a sound-wave generated by the implosion.

A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.  Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element. At some local point for a moment in time, there is enough energy for the water present to boil.  It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam ciculating in the water.  It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.

![Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs](/Users/aidanmorrison/submarines/EarlyCavitation.JPG){width=500px}

It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.  It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation. Larger bubbles result in lower frequencies of sound being emitted. Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.

![Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface](/Users/aidanmorrison/submarines/BoilingCavitation.JPG){width=500px}

Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages. It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.  Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.

Despite cavitation necessarily representing some energy being wasted generating unwanted noise in the water, some cavitation inevitably occurs around most propulsion systems operating at full power.  In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.  Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.  In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads [@shin2015]. Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.  It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely [@mollard2011], [@lewis1988].  These circumstances include the design of submarines and torpedoes.

In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.  However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.

Cavitation can occur in a number of different ways at different points and forms on or around the propeller.  It is common for cavitation to occur first near the out extremities of the propeller, where the blades are moving at the highest velocity relative to the water-flow.  Cavitation tends to spread across the back of the blades in a sheet (sheet cavitation), since the back side of the blades are generally the areas of most depressed pressure, though it can also occur in places on the back side of the blades, where the dynamics of the water moving around the blade can produce points of near the edges of significantly reduced pressure.   On surface vessels cavitation tends to occur most when the blades are closes to the surface, where the pressure is lowest.

![Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in [@mollard2011]](/Users/aidanmorrison/submarines/IncreasingCavitation.png){width=500px}

![Different types of cavitation can occur on different parts of the propulsion system as shown in [@carlton2007]](/Users/aidanmorrison/submarines/CavitationTypes.png){width=500px}

## Pressure, Thrust, and Momentum change
It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.  Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.  Generating thrust is consequently a key objective of any propulsion system.

Pressure is a measure of force divided by an area over which it is applied. Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.  In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some **net** pressure difference on different sides of an object.  For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.  Being generally backwards facing this acts to push the propeller forwards.  This pressure acts equally to push water backwards, as well as to push the propeller forwards.  Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.  It is the sum of all the pressure differences on the propeller that result in a net force.

Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.  In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.  Submerged jets represent a particular case in point.  A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.  When submerged, the backwards pressue appears dramatically reduced.  In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle. This effect represents one of the important distinctions between the jets which release their water below or above the water-line.

It is also an important law of physics that force or thrust is equal to the rate of change of momentum. Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.  Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust.  Circular movements necessarily are cancelling, and produce no thrust. It is useful to recall however that this principal is simply another complete and way of measuring the consequences of the net pressure differences discussed earlier.  The total rate of change of momentum in the water-column which is induced as a conseqence of the pressure differentials on all sides of the propeller must necessarily be equivalent to the total net force that is produced.  Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.

Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including [@lewis1988], [@carlton2007] and [@mollard2011].

## Flow separation
Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.  It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive. It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates [@li2013, @bakker2006, mcbride1979, wislicenus1986].

![Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)](/Users/aidanmorrison/submarines/SeparatedAerofoil.jpg){width=500px}

In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the pressure is increasing in the direction of movement, which means the flow is slowing down as it moves.  At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.  In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.

![Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in [@mollard2011]](/Users/aidanmorrison/submarines/FlowSeparation.png){width=500px}

In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow [@mollard2011 pp. 480].  If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.

# Basics of Ducted Propellers and Pumpjets
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.

## Propellers

Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.  A propeller uses a number of tilted blades attached to a central hub to sweep around disc in the water, and accelerate a column of water passing through the disc.  Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.  As such, it accelerates a column of water, which necessarily is contracted in circumference after the point of acceleration.

![Propellers generate thrust by accelerating a column of water, as shown in [@mollard2011 pp 247]](/Users/aidanmorrison/submarines/PropellerAction.png){width=500px}

A comprehensive description of the development of the propeller can be found in [@carlton2007].  Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.

Most propellers will have between three and seven blades.  In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.  The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.  The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake). Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.

![Propeller geometry including skew and rake, as shown in [@mollard2011 pp 262]](/Users/aidanmorrison/submarines/SkewRake.png){width=500px}

### Pitch
Perhaps one of the most important characteristics of a propeller is its pitch.  The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.  It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.  Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of propellers and propulsion systems.

![A blade section of a propeller traces a helical line around a cylinder as it is rotated [@carlton2007]](/Users/aidanmorrison/submarines/HelicalLine.png)

![The definition of pitch is the distance traveled by a blade section along a cylinder, as given in [@carlton2007]](/Users/aidanmorrison/submarines/PitchDefinition.png){width=500px}

The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in [@carlton2007 pp 35-37].

Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.  It is for this reason that a considerable number of vessels actually have variable pitch propellers.  Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.  Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.  This has advantages for ferries which undertake frequent docking manouvres in confined spaces [@MAN2017, @lewis1988].

![Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: [@MAN2017]](/Users/aidanmorrison/submarines/CPPropeller.png){width=500px}

### Pitch and the Advance Ratio
Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.

![A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN](/Users/aidanmorrison/submarines/OpenWater.png){width=500px}

This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.  However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case. The x-axis, $J$, also warrants a clear definition:

$$J = \frac{V_a}{nD}$$
Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed [@MAN2017 pp 15].  Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In [@lewis1988 pp 213] $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.

A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.  This is neatly defined and shown in a diagram in [@MAN2017].

![The slip ratio of a propeller. Image credit: [@MAN2017]](/Users/aidanmorrison/submarines/SlipRatio.png){width=500px}

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.

## Ducted Propellers
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. (Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted [@abdel2010].  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)

![The two types of ducted propeller, accelerating (left) and decelerating (right) as shown in [@mollard2011 pp 249]](/Users/aidanmorrison/submarines/AccelDecel.png){width=500px}

A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved. In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.  In the decellerating case, the available area contracts after the water has passed the impeller. In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal [@wislicenus1973].

![The streamlines of flow through different duct types [@lewis1988 pp 214]](/Users/aidanmorrison/submarines/DuctStreamlines.png){width=500px}

### The Accelerating Duct
Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.  In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.  Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.

The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward. This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.  An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.  As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips [@moulijn2015].
